---
title: "MovieLens Capstone Project for HarvardX PH125.9 Course"
author: "Rob Henderson"
date: '2022-07-07'
output: pdf_document:
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

#Introduction

This MovieLens Project is one of two projects for the Capstone portion of the HarvardX Data Science Course. Using information and techniques learned in the online course, as well as the provided dataset, the aim of this project is to develop and train a machine learning to Predict a Rating that will be given by a user for a movie.

The Residual Mean Square Error (RMSE) will be used to evaluate the accuracy of the Machine Learning Algorithm. For this project, the required criteria is an RMSE lower than 0.8775.

This report will offer an overview of the provided data, analysis and results of modeling that data and a conclusion.

##Initial Summary of Data Provided

First, we download and supplied dataset for the course, using the instructions provided.

```{r, echo=TRUE, message=FALSE}
##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################
# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1) # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

The dataset contains the variables userId, movieId, rating, timestamp, title and genres. These variables are the columns and the rows of the dataset correspond to each user's rating of a movie.

Looking at a summary of the dataset, we find that there are no missing values and that movies are rated between 0.5 and 5. We can also see that the mean rating given to movies is 3.512.

```{r, echo=TRUE, message=FALSE}
### Data Analysis ###

head(edx)
summary(edx)
```

Examining the dataset further, we can see that there are 10,677 unique movies, 797 unique genres and 69,878 unique users.
```{r, echo=TRUE, message=FALSE}
# Unique Movies, Genres and Users in the Dataset
edx %>%
summarize(unique_movies = n_distinct(movieId),
          unique_genres = n_distinct(genres),
            unique_users = n_distinct(userId))
```

Looking at a distribution of the ratings, we can see that users generally tend to rate movies between 3 and 5. It is also clear that users tend to rate movies as whole numbers, rather than half ratings.

```{r, echo=TRUE, message=FALSE}
# Distribution of Ratings
edx %>%
  ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.25, color = "black") +
  scale_y_continuous(breaks = c(seq(0, 3000000, 500000))) +
  xlab("Rating Given") +
  ylab("Number of Ratings") +
  ggtitle("Distribution of Ratings")
```

Further analysis shows that most users rate between 10 and 100 movies.

# Number of Ratings Per User
edx %>% 
  count(userId) %>%
  ggplot(aes(n)) +
  geom_histogram(binwidth = 0.15, color = "black") +
  scale_x_log10() +
  xlab("Number of Ratings") +
  ylab("Number of Users") +
  ggtitle("Number of Ratings Per User") 

##Analyzing the Data Further

# Mean Rating Given by Users

edx %>%
  group_by(userId) %>%
  filter(n() >= 100) %>%
  summarize(b_u = mean(rating)) %>%
  ggplot(aes(b_u)) +
  geom_histogram(bins = 30, color = "black") +
  xlab("Mean Rating Given") +
  ylab("Number of Users") +
  ggtitle("Mean Rating Given by Users") +
  scale_x_discrete(limits = c(seq(0.5,5,0.5))) 

#Predicting Using the Mean
mu <- mean(edx$rating)
mu

rmse_mean <- RMSE(validation$rating, mu)
rmse_mean

rmse_results = tibble(method = "Naive Analysis by Mean", RMSE = rmse_mean)
rmse_results

# Model Using the Movie Effect Bias Term

mu <- mean(edx$rating)

movie_averages <- edx %>%
  group_by(movieId) %>%
  summarise(b_i = mean(rating - mu))

predicted_ratings <- mu + validation %>%
  left_join(movie_averages, by='movieId') %>%
  pull(b_i)

rmse_model_movie_effects <- RMSE(predicted_ratings, validation$rating)
rmse_model_movie_effects
rmse_results <- bind_rows(rmse_results, 
                          data_frame(method="Movie Bias Model",
                                     RMSE = rmse_model_movie_effects))

rmse_results

## User Effects Model
## Simple model taking into account the user effects, b_u
user_averages <- edx %>%
  left_join(movie_averages, by="movieId") %>%
  group_by(userId) %>%
  summarise(b_u = mean(rating - mu - b_i))

predicted_ratings <- validation %>%
  left_join(movie_averages, by='movieId') %>%
  left_join(user_averages, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  pull(pred)

rmse_model_user_effects <- RMSE(predicted_ratings, validation$rating)
rmse_model_user_effects
rmse_results <- bind_rows(rmse_results, 
                          data_frame(method="Movie and User Bias Model",
                                     RMSE = rmse_model_user_effects))

rmse_results


## Regularization of Data

lambdas <- seq(0, 10, 0.25)
rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n() +l))
  
  b_u <- edx %>%
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- validation %>%
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
  return(RMSE(predicted_ratings, validation$rating))
  
})
rmse_regularization <- min(rmses)
rmse_regularization

# Plot RMSE against Lambdas to find optimal lambda
qplot(lambdas, rmses)
lambda <- lambdas[which.min(rmses)]
lambda

rmse_results <- bind_rows(rmse_results, 
                          data_frame(method="Regularization",
                                     RMSE = rmse_model_user_effects))

## Final results
rmse_results